A Meta Analysis of the Bacterial Microbiome and Human Obesity

### 2016-02-12 code/Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
manuscript update 2

*Sequence Analysis Pipeline*
<br><br>
All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive.  A total of 7/11 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP.  A total of 3/11 [@goodrich_human_2014; @yatsunenko_human_2012] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP.  The final data set [@arumugam_enterotypes_2011] that was used took advantage of the Metaphlan2 package [@truong_metaphlan2_2015] to generate the necessary data for this analysis.  A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1 & S2].  For a few of the studies after sequence processing there was a large decrease in the number of samples available for analysis [Table 1].  A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/marcsze/MOM-Analysis.

*Data Analysis*
<br><br>
All analysis was performed using R version 3.2.2 and R studio version 0.99.491.  The overall analysis was split into three general categories.  The first category involved following the same method employed by Finucane et al. and Walters et al. [-@finucane_taxonomic_2014; -@walters_meta-analyses_2014] in which each study was re-analyzed seperately for associations with BMI.  The main associations with BMI that were explored in each seperate data set were Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR).  A two-way T-test was performed for comparison for non-obese versus obese.  An ANOVA was used to investigate all BMI categories (if available) with a Tukey post hoc test. Third, a Pearson test for linear correlation between the variables was used.  Finally, Random forest analysis to identify the best variables for classification of obese and non-obese was performed to look if there were similarities between data sets that we may have missed.  The R package AUCRF (v1.1) was used for this.
<br><br>
The second category involved pooling all the data from all the studies together to asses whether the combined direction of all the studies increased an indivdiual's RR for being obese.  The two measurements that were investigated in detail for this part were Shannon diversity and the B/F ratio.  To generate the RR the median value of each study was taken and the number of those that were non-obese or obese in each group was recorded and a Fisher exact-test was performed.  The relative risk from the the Fisher exact-test was obtained using the epiR (v0.9-6.9) package in R.  To obtain the pooled RR and confidence intervals (CI) the metafor (v1.9-8) package available for the R statistics software was utilized.  If the pooled analysis CI does not touch or cross 1.00 the pooled RR had a P-value < 0.05. 
<br><br>
The third and final category involved using mostly Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.  Using the Zscore values a two-way t-test was used to investigate if there was a difference between non-obese and obese individuals.  Differences in BMI categories (if data was available) was investigated using ANOVA with a tukey post hoc test.  Linear correlations between these variables and BMI were explored using the Pearson test.  Where appropriate the Bonferroni correction was applied for multiple comparisons.  In all other cases a P-value under 0.05 was considered to be significant.

@@

*******
**Table 1:Summary Information of Studies used in Meta-analysis.**     
![Table 1](C:\Users\marc\Desktop\obesity2\SummaryTableStudies.tif)

>>>>

*Sequence Analysis Pipeline*
<br><br>
All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive.  A total of 7/11 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @nam_comparative_2011; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP.  A total of 3/11 [@goodrich_human_2014; @yatsunenko_human_2012] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP.  The final data set [@arumugam_enterotypes_2011] that was used took advantage of the Metaphlan2 package [@truong_metaphlan2_2015] to generate the necessary data for this analysis.  A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1 & S2].  For a few of the studies after sequence processing there was a large decrease in the number of samples available for analysis [Table 1].  A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/blob/master/code/sequenceProcessing.md.

*Data Analysis*
<br><br>
All analysis was performed using R version 3.2.2 and R studio version 0.99.491.  The overall analysis was split into three general categories.  The first category involved following the same method employed by Finucane et al. and Walters et al. [-@finucane_taxonomic_2014; -@walters_meta-analyses_2014] in which each study was re-analyzed seperately for associations with BMI.  The main associations with BMI that were explored in each seperate data set were Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR).  A two-way T-test was performed for comparison for non-obese versus obese.  An ANOVA was used to investigate all BMI categories (if available) with a Tukey post hoc test. Third, a Pearson test for linear correlation between the variables was used.  Finally, Random forest analysis to identify the best variables for classification of obese and non-obese was performed to look if there were similarities between data sets that we may have missed.  The R package AUCRF (v1.1) was used for this.
<br><br>
The second category involved pooling all the data from all the studies together to asses whether the combined direction of all the studies increased an indivdiual's RR for being obese.  The two measurements that were investigated in detail for this part were Shannon diversity and the B/F ratio.  To generate the RR the median value of each study was taken and the number of those that were non-obese or obese in each group was recorded and a Fisher exact-test was performed.  The relative risk from the the Fisher exact-test was obtained using the epiR (v0.9-6.9) package in R.  To obtain the pooled RR and confidence intervals (CI) the metafor (v1.9-8) package available for the R statistics software was utilized.  If the pooled analysis CI does not touch or cross 1.00 the pooled RR had a P-value < 0.05. 
<br><br>
The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.  Using the Zscore values a two-way t-test was used to investigate if there was a difference between non-obese and obese individuals.  Differences in BMI categories (if data was available) was investigated using ANOVA with a tukey post hoc test.  Linear correlations between these variables and BMI were explored using the Pearson test.  Where appropriate the Bonferroni correction was applied for multiple comparisons.  In all other cases a P-value under 0.05 was considered to be significant.

@@

*******
```{r data_loading, echo=FALSE, include=FALSE}

```

**Table 1:Summary Information of Studies used in Meta-analysis.**     
```{r Table_1, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)

Study <- c("Baxter", "Ross", "Goodrich", "Escobar", "Zupancic", "Yatsunenko", "HMP", "Nam", "Wu", "Arumugam", "Turnbaugh")

Year <- c(2016, 2015, 2014, 2014, 2012, 2012, 2011, 2011, 2011, 2010, 2009)

Total <- c(172, 63, 525, 30, 227, 531, 256, 18, 65, 85, 154)

Processed <- c(172, 63, 507, 30, 207, "In Progress", 256, 18, 58, 85, 146)

table1 <- cbind(Study, Year, Total, Processed)

colnames(table1) <- c("Study", "Year", "Total n", "n After Sequence Processing")

kable(table1, format = "markdown", align = 'c', padding = 2)

```
